---
title: "EDA"
author: "Yiying Wu"
date: "2023-10-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## data import
`lubridate::floor_date`: round date times down
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())

```
##Initial numeric explorations

```{r}
weather_df |> 
  ggplot(aes(x = prcp)) + 
  geom_histogram()
```
heavy skewed

Here are data outliers
```{r}
weather_df |> 
  filter(prcp >= 1000)
```

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) + 
  geom_point(alpha = .75)
```

## `group by`
```{r}
weather_df |>
  group_by(name, month)
```
it is sometimes necessary to remove groups using `ungroup()`.

## Counting things
`summarize`
```{r}
weather_df |>
  group_by(month) |>
  summarize(n_obs = n())
```
group by more than one variable
```{r}
weather_df |>
  group_by(name, month) |>
  summarize(n_obs = n())
```

```{r}
weather_df |>
  count(name, month) |>
  pivot_wider(
    names_from=name,
    values_from=n
  )
```

## General summaries
`summarize()`
```{r}
weather_df |>
  group_by(name,month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prec = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax),
    sd_tmax = sd(tmax))
```
create a plot based on the monthly summary
```{r}
weather_df |>
  group_by(name, month) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```
`knitr::kable(digits=2)`: put the result into tables
```{r}
weather_df |>
  group_by(name, month) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE))|>
  pivot_wider(
    names_from=name,
    values_from=mean_tmax
  )|>
  knitr::kable(digits=2)
```

## Grouped mutate
```{r}
weather_df |>
  group_by(name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax) |> 
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point() 
```

